A Plasma Torus Around a Young Low-Mass Star

### 2025-03-28 paper/ms.tex
update ms: JSS comments v0

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
coherent sinusoidal Balmer emission at twice to four times the
star's equatorial velocity, directly demonstrating for the first
time the existence of corotating circumstellar plasma around a CPV.
Given that long-lived condensations of cool ($10^4$ K) plasma can
persist in the hot ($10^6$ K) coronae of stars with a wide range of
masses \cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can sustain an optically thick component; the origin
of the optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/f1.pdf}
  \captionsetup{labelformat=empty}
  \caption{{\bf Figure 1 (Movie):  TIC~141146667 is a complex periodic
  variable (CPV).}  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_20250116.mp4}{available here},
  covers a baseline of 5{,}784 cycles irregularly sampled over three
  years.  The TESS light curve is phased to the \periodhr\ hour period
  in groups of a few cycles per frame.  This is the period both of
  stellar rotation, and (we hypothesize) of corotating clumps of
  circumstellar material.  Raw data acquired with two minute sampling
  are in gray; black averages to 100 points per cycle.  Similar to
  other stars of this class, the sharp photometric features persist
  for tens to thousands of rotational cycles. }
  \label{fig:lc}
\end{figure}

@@

In this study, we present the first spectroscopic detection of
corotating clumps of cool plasma around a CPV.  We identified
TIC~141146667 in previous work \cite{Bouma2024} by searching the TESS
two-minute data \cite{Ricker2015} for stars showing highly structured,
periodic light curves.  We selected the star for spectroscopy because
its brightness and rotation rate enable measuring variability in its line
profile.  We observed it for five hours on
17~February~2024 (UT) using the High Resolution Echelle Spectrometer
(HIRES \cite{vogt_hires_1994}) on the 10\,m Keck I telescope.  TESS
observed the star over a four week interval from 30~January~2024 to
26~February~2024 with a duty cycle of 77\%.  During the HIRES
observations, TESS was performing a data downlink; TESS data
collection resumed 12 hours (three rotation cycles) after the spectra
were acquired.  Extended Data Figure~\ref{fig:fulllc} shows the
detailed photometric behavior of the star near the epoch of
observation; the general photometric morphology is similar before and
after the data gap.

@@

Spectra of magnetically-active, rapidly rotating stars with a wide range
of masses have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} and
transient absorption features
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for
such spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar corona
that can last days to weeks \cite{VialEngvold2015}.  These
condensations fall back to the Sun's surface because gravity is
stronger than any magnetic tension or centrifugal force capable of
sustaining them.  However for stars with magnetospheric radii $R_{\rm
m}$ that exceed their corotation radii $R_{\rm c}$, the effective
potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to accumulate over much
longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit, nor be optically thick.
For instance, the H$\alpha$ emission clump at 3.9\,$R_\star$ shows no 
photometric signal in the February 2024 photometry.

@@

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds that
accumulate into warped plasma tori \cite{Townsend2005,Townsend2008}.
These tori tend to have dense antipodal accumulations of plasma
sculpted by tilted-dipole magnetic fields, and the transits of these
clumps produce broadband photometric variability through bound-free
scattering \cite{Townsend2005} and Thomson scattering
\cite{Berry2022}.  
The plasma accumulates at antipodes 180$^\circ$ apart because the
deepest local minima in the effective potential exist along the line
of intersection between the rotational and magnetic equators (see
Equation~22 of \cite{Townsend2005}).
For $\sigma$~Ori~E and almost
all of its analogs, the result is light curves that appear ``simple'',
resembling those of eclipsing binaries, and time-dynamic H$\alpha$
spectra similar to those in Figure~\ref{fig:spec}
\cite{Townsend2005,Townsend2008}.  The two known exceptions, HD~37776
and HD~64740, show complex light curves resembling CPVs
\cite{Mikulasek2020,Bouma2024} and have spectropolarimetric magnetic
field maps indicating strong contributions from higher-order magnetic
moments \cite{Kochukhov2011,Shultz2018}.  There are two implications:
first, the complexity of CPVs may be a direct consequence of magnetic
fields with highly multipolar contributions.  Second, CPVs can be a
source of astrophysical false positives in photometric classification
of eclipsing binaries and transiting exoplanets around young pre-main
sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

Figure~\ref{fig:fulllc} shows the TESS Sector~75 data acquired near
the epoch of spectroscopy.  The gap in coverage from BTJD
3356.0~-~3358.5 occurred because the Earth was within 25$^\circ$ of
the TESS camera's boresight, yielding high levels of scattered light;
this gap included the Keck/HIRES observation epoch (green bar).  From
BTJD 3359.4~-~3362.0, the Moon then passed within 25$^\circ$ of the
camera's boresight.  Based on the observed level of scattered light in
the optimal TIC~141146667 aperture, we manually masked out times from
3359.40~-~3360.13, and judged the remainder of the data during the
lunar approach to be usable.  Small data gaps from BTJD
3353.55~-~3353.77 and from BTJD 3360.12~-~3360.33 were caused by data
downlinks at the spacecraft's perigee and apogee, respectively.

Figure~\ref{fig:fulllc} shows that a large, complex eclipse
feature was present before and after the HIRES data were
acquired.  From BTJD 3352~-~3356, this eclipse had two sharp local
minima;  the local minimum at $\phi$$\approx$0.8 disappeared following
the flare at BTJD 3353, yielding an eclipse more closely resembling a
long ``V'' than a ``W''.  Similar state changes have previously been
noted and discussed \cite{Stauffer2017,Bouma2024}.  The photometric
shape therefore evolved during the twelve cycles spanning the
3356~-~3358 gap, since the average shape from 3358-3363 more
closely resembles the initial ``W''.

@@

{\it Effective temperature, radius, mass, and spectral
classification}---We adopted the star's effective temperature and
radius measured using the spectral energy distribution (SED) fitting
procedure described by \cite{Bouma2024}.  To summarize, this approach
used \texttt{astroARIADNE} \cite{Vines2022} with the BT-Settl stellar
atmosphere models \cite{Allard2012}, assuming the \cite{Asplund2009}
solar abundances and the \cite{Barber2006} water line lists.  This
approach fitted for the stellar effective temperature, radius,
reddening, surface gravity, metallicity, and distance by comparing the
measured broadband magnitudes against pre-computed model grids.
Specifically, we performed the fit using the broadband magnitudes from
Gaia DR2, APASS, 2MASS, SDSS, and WISE $W1$ and $W2$.  The resulting
best-fit SED is shown in Figure~\ref{fig:sed}.  This method has the
most constraining power for the star's effective temperature (2972
$\pm$ 40\,K) and radius (0.42 $\pm$ 0.02\,$R_\odot$); the surface
gravity, metallicity, and reddening are only weakly constrained.  We
determined the star's spectral type to be M5.5Ve by visually comparing
the HIRES spectra against the photometric standards tabulated by
\cite{Bochanski2007}.   This result agrees with the effective
temperature found from the SED fitting \cite{Pecaut2013}.  We measured
the equivalent width of the H$\alpha$ line by fitting a range of
models to the time-averaged line profile shown in
Figure~\ref{fig:spec}, selecting the model that minimized the Bayesian
information criterion, and numerically integrating this best fit model.  
We found a sum of two Gaussians to be the preferred model; our quoted
result, EW$_{{\rm H}\alpha}$=7.2$\pm$0.2\,\AA, comes from numerically
integrating within $|\Delta v/v_{\rm eq}|<1$.  Integrating over the
entire line profile, including the broad wings, would yield EW$_{{\rm
H}\alpha}$=10.2$\pm$0.3\,\AA. Either value would classify the star as
a weak-lined T Tauri \cite{Briceno2019}.

@@

{\it Age: Isochrones}---The color and absolute magnitude of
TIC~141146667 suggest that it is a pre-main sequence M dwarf, similar to
all other known CPVs \cite{Stauffer2017,Stauffer2021,Bouma2024}.  The
star's proximity ($d$=58\,pc) and high galactic latitude
($b$=$+$53$^\circ$) yield negligible interstellar reddening along the
line of sight \cite{Green2019}.  Figure~\ref{fig:camd} shows the
location of TIC~141146667 in the color--absolute magnitude diagram
(CAMD) relative to young stellar populations including Upper Scorpius
(USco), IC~2602, and the Pleiades.  To make this diagram, we adopted
the USco members in the $\delta$~Sco and $\sigma$~Sco sub-associations
from \cite{Ratzenbock2023}, and the IC~2602 and Pleiades members from
\cite{Hunt2024}.  We assumed an average $V$-band extinction $A_{\rm
V}$=$\{0.12, 0.11, 0.10\}\,{\rm mag}$ for USco \cite{Pecaut2016},
IC~2602 \cite{Hunt2024}, and the Pleiades \cite{Hunt2024}
respectively, and ages of 8\,Myr \cite{Ratzenbock2023}, 40\,Myr
\cite{Randich2018}, and 112\,Myr \cite{Dahm2015} for each respective
cluster.  We dereddened the photometry using the extinction
coefficients $k_X\equiv A_X/A_0$ tabulated in
\cite{GaiaCollaboration2018}, assuming that $A_0 = 3.1 E(B-V)$.

Figure~\ref{fig:camd} shows that TIC~141146667 falls between the USco
and Pleiades sequences, and approximately overlaps IC~2602.  However,
the precision of the implied age is set by the intrinsic scatter of
these calibration sequences; the most luminous stars in the Pleiades of
the same color have a similar absolute magnitude as TIC~141146667.
Previous work \cite{Stauffer2021} has also noted that in the Gaia
passbands, CPVs tend to be photometrically redder and more luminous than
single stars in any given cluster, similar to other rapid rotators.
While this effect complicates any attempt at age inference based on the
Gaia photometry, it suggests that the Pleiades may be a better
comparison population than IC~2602.  We take the star's
location in the color--absolute magnitude diagram to suggest age bounds
$t_{\rm CAMD}$$\sim$30-150\,Myr.

@@

Despite uncertainties in the details, what can be stated with confidence
is that lithium is not abundant in the spectrum of TIC~141146667.
Figure~\ref{fig:liew_population} compares our upper limit against the
equivalent width measurements reported by \cite{Jeffries2023}
based on the Gaia-ESO spectroscopic survey; Figure~\ref{fig:hirescuts}
shows the associated raw spectra.  If the star were $\lesssim$20\,Myr
old, at its temperature we would expect to see lithium in abundance
($>$400\,m\AA).  Since we do not, we can set an empirical bound on the
lithium-derived age of $t_{\rm Li,emp}$$\gtrsim$20\,Myr.  The
\cite{Feiden2016} lithium isochrones provide a point for theoretical
comparison, and suggest that since $M_{\rm K}$$\approx$6.67\,mag,
$t_{\rm Li,th}$$\gtrsim$35\,Myr is the theoretical age at which
complete depletion occurs in a star with this luminosity (see
e.g.~Figure 7 from \cite{Wood2023}).

@@

{\it Age: Summary}---The main indicators for the youth of
TIC~141146667 are {\it i)} that it is a complex periodic variable, and
{\it ii)} that it is 1.5 magnitudes brighter (four times more
luminous) than main sequence stars of the same color, while showing no
indicators for binarity.  The first item constitutes evidence for
youth since all known CPVs are $\lesssim$200\,Myr \cite{Bouma2024}.
The reference sequences for the Pleiades (112\,Myr, \cite{Dahm2015})
show a few stars of equal luminosity and the same temperature,
suggesting a photometric isochronal age upper limit
$\lesssim$150\,Myr.  The weak lithium absorption suggests an age of at
least 20\,Myr based on an empirical comparison using Gaia-ESO spectra,
or at least 35\,Myr based on the \cite{Feiden2016} isochrones.  These
considerations yield our adopted age range of 150\,Myr.

@@

Figure~\ref{fig:hirescuts} shows a few regions of interest in the
HIRES spectra, which cover 3650-7960\,\AA.  Higher order Balmer lines
including H$\gamma$ and H$\delta$ ($n=5\rightarrow2$ and
$n=6\rightarrow2$) do show variability outside the line core.
However, this variability is not clearly periodic in the same manner
as the emission seen in H$\alpha$.  This could be because there are
insufficient hydrogen ions in the relevant excited states, or because
the spectra have lower precision in these bluer regions.  The Ca HK
doublet is also detected in emission, while the continuum near it is
not.  Chromospheric emission from the magnesium triplet is also
detected, but these lines are sufficiently blended to be rendered
unusable.

Figure~\ref{fig:hirescuts} provides a novel view on the sudden blue
excess that appeared at $\phi$$\approx$0.2 in H$\alpha$, H$\gamma$,
H$\delta$, and He 5876\,\AA.  At the same epoch, only H$\alpha$
clearly shows a resolved red excess as well, although H$\gamma$ and
H$\delta$ are broadened at this time.  Given the sudden rise of this
emission in $<$15\,minutes, this sudden blue excess might suggest a
more sudden flow, rather than a stable, periodic component.  A stellar
flare could be a plausible explanation for such a sudden rise.
However this idea seems incompatible with the sinusoidal emission seen
from $\phi$$\approx$0.5-1.0, and with the fact that flares often
excite iron lines in the blue HIRES orders, which are not observed.
More time-series spectroscopy would be needed to provide a definitive
clarification for this type of non-periodic variability.

Finally, the Li\,\textsc{I} 6708\,\AA\ doublet, which shows no obvious
absorption, as well as the broad K\,\textsc{I} 7699\,\AA\ resonance
line are both shown in Figure~\ref{fig:hirescuts}.  In the latter,
narrow telluric absorption features overlap the blue wing of the line.
Neither of these regions shows any notable variability.

@@

Regarding the upper limit, Figure~\ref{fig:sed} shows the SED.  While
AllWISE \cite{Cutri2014} yielded a confident W3 detection
(9.8$\sigma$), the W4 extraction yielded only a marginal indication
(1.7$\sigma$) of detectable flux.  Similar to other CPVs
\cite{Stauffer2017,Bouma2024}, the photometric uncertainties from WISE
W1 and W2 allow at most a $\lesssim$2\% excess at 3-5\,$\mu$m relative
to the stellar photosphere, and a $\lesssim$5\% excess at 10\,$\mu$m
(W3).  To estimate the implied mass bound, we assume a dust
temperature $T_{\rm d}$=1500\,K, typical for dust near the star (see
\cite{Zhan2019} for discussion regarding dust sublimation).  We then
treat emission from the dust and star as Planck functions, and require
$L_{\rm d} < f L_\star$, where the factor $f$ is set by the
photometric precision of WISE and $L_{\rm d}$ is the bolometric dust
luminosity.  Given the reported uncertainties, we numerically find
$f<6\cdot$10$^{-3}$.  From the Stefan-Boltzmann law we can then write
$A_{\rm d} < f (T_\star / T_{\rm d})^4 Q_{\rm em}^{-1} (4\pi
R_\star)^2$, for $A_{\rm d}$ the total emitting surface area of the
dust, and $Q_{\rm em}$ an emission efficiency parameter.  Converting
this constraint to a dust mass requires an assumption regarding the
grain properties.  We assume a grain density $\rho_{\rm
d}$=3\,g\,cm$^{-3}$ typical for silicate grains, a fixed grain size
$a$=1\,$\mu$m, and no self-absorption.  This enables the assumption
that $A_{\rm d} = N \pi a^2$, for $N$ the total number of dust grains.
This in turn yields an upper limit on the dust mass of
\begin{equation}
  M_{\rm dust} \lesssim 4 \cdot 10^{17}\, {\rm g}\ 
  \left( \frac{f}{6\cdot10^{-3}} \right)
  \left( \frac{T_\star}{3000\,{\rm K}} \frac{1500\,{\rm K}}{T_{\rm d}} \right)^4
  \left( \frac{Q_{\rm em}}{1} \right)^{-1}
  \left( \frac{R_\star}{0.4\,R_\odot} \right)^2
  \left( \frac{a}{1\,\mathrm{\mu m}} \right)
  \left( \frac{\rho_{\rm d}}{3\,\mathrm{g\,cm^{-3}}} \right).
\end{equation}

>>>>

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  However, the
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
coherent sinusoidal Balmer emission at twice to four times the
star's equatorial velocity, directly demonstrating for the first
time the existence of corotating circumstellar plasma around a CPV.
Given that long-lived condensations of cool ($10^4$ K) plasma can
persist in the hot ($10^6$ K) coronae of stars with a wide range of
masses \cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can sustain an optically thick component; the origin
of the optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/f1.pdf}
  \captionsetup{labelformat=empty}
  \caption{{\bf Figure 1 (Movie):  TIC~141146667 is a complex periodic
  variable (CPV).}  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_20250116.mp4}{available here},
  covers a baseline of 5{,}784 cycles irregularly sampled over three
  years.  The TESS light curve is phased to the \periodhr\ hour period
  in groups of three cycles per frame.  This is the period both of
  stellar rotation, and (we hypothesize) of corotating clumps of
  circumstellar material.  Raw data acquired with two minute sampling
  are in gray; black averages to 100 points per cycle.  Similar to
  other stars of this class, the sharp photometric features persist
  for tens to thousands of rotational cycles. }
  \label{fig:lc}
\end{figure}

@@

In this study, we present the first spectroscopic detection of
corotating clumps of cool plasma around a CPV.  We identified
TIC~141146667 in previous work \cite{Bouma2024} by searching the TESS
two-minute data \cite{Ricker2015} for stars showing highly structured,
periodic light curves.  We selected the star for spectroscopy because
its brightness and rotation rate enable efficiently measuring
variability in its line profile.  We observed it for five hours on
17~February~2024 (UT) using the High Resolution Echelle Spectrometer
(HIRES \cite{vogt_hires_1994}) on the 10\,m Keck I telescope.  TESS
observed the star over a four week interval from 30~January~2024 to
26~February~2024 with a duty cycle of 77\%.  During the HIRES
observations, TESS was performing a data downlink; TESS data
collection resumed 12 hours (three rotation cycles) after the spectra
were acquired.  Extended Data Figure~\ref{fig:fulllc} shows the
detailed photometric behavior of the star near the epoch of
observation; the general photometric morphology is similar before and
after the data gap.

@@

Spectra of magnetically-active, rapidly rotating stars with a wide range
of masses have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} and
transient absorption features
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for
such spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar corona
that can last days to weeks \cite{VialEngvold2015}.  These
condensations fall back to the Sun's surface because gravity is
stronger than any magnetic tension or centrifugal force capable of
sustaining them.  However for stars with magnetospheric radii $R_{\rm
m}$ that exceed their corotation radii $R_{\rm c}$, the effective
potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to accumulate over much
longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit nor be opaque in
broadband optical light.
For instance, the H$\alpha$ emission clump at 3.9\,$R_\star$ shows no 
photometric signal in the February 2024 photometry.

@@

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds that
accumulate into warped plasma tori \cite{Townsend2005,Townsend2008}.
These tori tend to have dense antipodal accumulations of plasma
sculpted by tilted-dipole magnetic fields, and the transits of these
clumps produce broadband photometric variability through bound-free
scattering \cite{Townsend2005} and Thomson scattering
\cite{Berry2022}.  
The plasma accumulates at antipodes 180$^\circ$ apart because the
deepest local minima in the effective potential exist along the line
of intersection between the rotational and magnetic equators (see
Equation~22 of \cite{Townsend2005}).
For $\sigma$~Ori~E and almost
all of its analogs, the result is ``simple'' light curves that 
resemble those of eclipsing binaries, and time-dynamic H$\alpha$
spectra similar to those in Figure~\ref{fig:spec}
\cite{Townsend2005,Townsend2008}.  The two known exceptions, HD~37776
and HD~64740, show complex light curves resembling CPVs
\cite{Mikulasek2020,Bouma2024} and have spectropolarimetric magnetic
field maps indicating strong contributions from higher-order magnetic
moments \cite{Kochukhov2011,Shultz2018}.  There are two implications:
first, the complexity of CPVs may be a direct consequence of magnetic
fields with highly multipolar contributions.  Second, CPVs can be a
source of astrophysical false positives in photometric classification
of eclipsing binaries and transiting exoplanets around young pre-main
sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

Extended Data Figure~\ref{fig:fulllc} shows the TESS Sector~75 data acquired near
the epoch of spectroscopy.  The gap in coverage from BTJD
3356.0~-~3358.5 occurred because the Earth was within 25$^\circ$ of
the TESS camera's boresight, yielding high levels of scattered light;
this gap included the Keck/HIRES observation epoch (green bar).  From
BTJD 3359.4~-~3362.0, the Moon then passed within 25$^\circ$ of the
camera's boresight.  Based on the observed level of scattered light in
the optimal TIC~141146667 aperture, we manually masked out times from
3359.40~-~3360.13, and judged the remainder of the data during the
lunar approach to be usable.  Small data gaps from BTJD
3353.55~-~3353.77 and from BTJD 3360.12~-~3360.33 were caused by data
downlinks at the spacecraft's perigee and apogee, respectively.

Extended Data Figure~\ref{fig:fulllc} shows that a large, complex eclipse
feature was present before and after the HIRES data were
acquired.  From BTJD 3352~-~3356, this eclipse had two sharp local
minima;  the local minimum at $\phi$$\approx$0.8 disappeared following
the flare at BTJD 3353, yielding an eclipse more closely resembling a
long ``V'' than a ``W''.  Similar state changes have previously been
noted and discussed \cite{Stauffer2017,Bouma2024}.  The photometric
shape therefore evolved during the twelve cycles spanning the
3356~-~3358 gap, since the average shape from 3358-3363 more
closely resembles the initial ``W''.

@@

{\it Effective temperature, radius, mass, and spectral
classification}---We adopted the star's effective temperature and
radius measured using the spectral energy distribution (SED) fitting
procedure described by \cite{Bouma2024}.  To summarize, this approach
used \texttt{astroARIADNE} \cite{Vines2022} with the BT-Settl stellar
atmosphere models \cite{Allard2012}, assuming the \cite{Asplund2009}
solar abundances and the \cite{Barber2006} water line lists.  This
approach fitted for the stellar effective temperature, radius,
reddening, surface gravity, metallicity, and distance by comparing the
measured broadband magnitudes against pre-computed model grids.
Specifically, we performed the fit using the broadband magnitudes from
Gaia DR2, APASS, 2MASS, SDSS, and WISE $W1$ and $W2$.  The resulting
best-fit SED is shown in Extended Data Figure Figure~\ref{fig:sed}.  This method has the
most constraining power for the star's effective temperature (2972
$\pm$ 40\,K) and radius (0.42 $\pm$ 0.02\,$R_\odot$); the surface
gravity, metallicity, and reddening are only weakly constrained.  We
determined the star's spectral type to be M5.5Ve by visually comparing
the HIRES spectra against the photometric standards tabulated by
\cite{Bochanski2007}.   This result agrees with the effective
temperature found from the SED fitting \cite{Pecaut2013}.  We measured
the equivalent width of the H$\alpha$ line by fitting a range of
models to the time-averaged line profile shown in
Figure~\ref{fig:spec}, selecting the model that minimized the Bayesian
information criterion, and numerically integrating this best fit model.  
We found a sum of two Gaussians to be the preferred model; our quoted
result, EW$_{{\rm H}\alpha}$=7.2$\pm$0.2\,\AA, comes from numerically
integrating within $|\Delta v/v_{\rm eq}|<1$.  Integrating over the
entire line profile, including the broad wings, would yield EW$_{{\rm
H}\alpha}$=10.2$\pm$0.3\,\AA. Either value would classify the star as
a weak-lined T Tauri \cite{Briceno2019}.

@@

{\it Age: Isochrones}---The color and absolute magnitude of
TIC~141146667 suggest that it is a pre-main sequence M dwarf, similar to
all other known CPVs \cite{Stauffer2017,Stauffer2021,Bouma2024}.  The
star's proximity ($d$=58\,pc) and high galactic latitude
($b$=$+$53$^\circ$) yield negligible interstellar reddening along the
line of sight \cite{Green2019}.  Extended Data Figure~\ref{fig:camd} shows the
location of TIC~141146667 in the color--absolute magnitude diagram
(CAMD) relative to young stellar populations including Upper Scorpius
(USco), IC~2602, and the Pleiades.  To make this diagram, we adopted
the USco members in the $\delta$~Sco and $\sigma$~Sco sub-associations
from \cite{Ratzenbock2023}, and the IC~2602 and Pleiades members from
\cite{Hunt2024}.  We assumed an average $V$-band extinction $A_{\rm
V}$=$\{0.12, 0.11, 0.10\}\,{\rm mag}$ for USco \cite{Pecaut2016},
IC~2602 \cite{Hunt2024}, and the Pleiades \cite{Hunt2024}
respectively, and ages of 8\,Myr \cite{Ratzenbock2023}, 40\,Myr
\cite{Randich2018}, and 112\,Myr \cite{Dahm2015} for each respective
cluster.  We dereddened the photometry using the extinction
coefficients $k_X\equiv A_X/A_0$ tabulated in
\cite{GaiaCollaboration2018}, assuming that $A_0 = 3.1 E(B-V)$.

Extended Data Figure~\ref{fig:camd} shows that TIC~141146667 falls between the USco
and Pleiades sequences, and approximately overlaps IC~2602.  However,
the precision of the implied age is set by the intrinsic scatter of
these calibration sequences; the most luminous stars in the Pleiades of
the same color have a similar absolute magnitude as TIC~141146667.
Previous work \cite{Stauffer2021} has also noted that in the Gaia
passbands, CPVs tend to be photometrically redder and more luminous than
single stars in any given cluster, similar to other rapid rotators.
While this effect complicates any attempt at age inference based on the
Gaia photometry, it suggests that the Pleiades may be a better
comparison population than IC~2602.  We take the star's
location in the color--absolute magnitude diagram to suggest age bounds
$t_{\rm CAMD}$$\sim$30-150\,Myr.

@@

Despite uncertainties in the details, what can be stated with confidence
is that lithium is not abundant in the spectrum of TIC~141146667.
Extended Data Figure~\ref{fig:liew_population} compares our upper limit against the
equivalent width measurements reported by \cite{Jeffries2023}
based on the Gaia-ESO spectroscopic survey; Extended Data Figure~\ref{fig:hirescuts}
shows the associated raw spectra.  If the star were $\lesssim$20\,Myr
old, at its temperature we would expect to see lithium in abundance
($>$400\,m\AA).  Since we do not, we can set an empirical bound on the
lithium-derived age of $t_{\rm Li,emp}$$\gtrsim$20\,Myr.  The
\cite{Feiden2016} lithium isochrones provide a point for theoretical
comparison, and suggest that since $M_{\rm K}$$\approx$6.67\,mag,
$t_{\rm Li,th}$$\gtrsim$35\,Myr is the theoretical age at which
complete depletion occurs in a star with this luminosity (see
e.g.~Figure 7 from \cite{Wood2023}).

@@

{\it Age: Summary}---The main indicators for the youth of
TIC~141146667 are {\it i)} that it is a complex periodic variable, and
{\it ii)} that it is 1.5 magnitudes brighter (four times more
luminous) than main sequence stars of the same color, while showing no
indicators for binarity.  Being a CPV suggests that the star is young
because a previous CPV search unbiased in age found 90\% of its
detections to be in $\lesssim$200\,Myr old clusters \cite{Bouma2024};
the remaining 10\% could not be associated with any coeval
population.
Similarly, studies of rotation in $\lesssim$100\,Myr clusters serendipitously found
$\approx$50-100 examples of the class
\cite{Rebull2016,Stauffer2017,Stauffer2018,Rebull2018,Zhan2019,Rebull2020,Stauffer2021,Rebull2022,Popinchalk2023},
whereas analogous studies of Praesepe and the Hyades did not report
any evidence for CPVs in a set of approximately one thousand
$\approx$700\,Myr stars \cite{Rebull2017,Douglas2019,Rampalli2021}.
Regarding the isochronal age constraint, the Pleiades (112\,Myr,
\cite{Dahm2015}) shows a few stars of equal luminosity and the same
temperature, suggesting a photometric isochronal age upper limit
$\lesssim$150\,Myr.  The weak lithium absorption suggests an age of at
least 20\,Myr based on an empirical comparison using Gaia-ESO spectra,
or at least 35\,Myr based on the \cite{Feiden2016} isochrones.  These
considerations yield our adopted age range of 35-150\,Myr.

@@

Extended Data Figure~\ref{fig:hirescuts} shows a few regions of interest in the
HIRES spectra, which cover 3650-7960\,\AA.  Higher order Balmer lines
including H$\gamma$ and H$\delta$ ($n=5\rightarrow2$ and
$n=6\rightarrow2$) do show variability outside the line core.
However, this variability is not clearly periodic in the same manner
as the emission seen in H$\alpha$.  This could be because there are
insufficient hydrogen ions in the relevant excited states, or because
the spectra have lower precision in these bluer regions.  The Ca HK
doublet is also detected in emission, while the continuum near it is
not.  Chromospheric emission from the magnesium triplet is also
detected, but these lines are sufficiently blended to be rendered
unusable.

Extended Data Figure~\ref{fig:hirescuts} provides a novel view on the sudden blue
excess that appeared at $\phi$$\approx$0.2 in H$\alpha$, H$\gamma$,
H$\delta$, and He 5876\,\AA.  At the same epoch, only H$\alpha$
clearly shows a resolved red excess as well, although H$\gamma$ and
H$\delta$ are broadened at this time.  Given the sudden rise of this
emission in $<$15\,minutes, this sudden blue excess might suggest a
more sudden flow, rather than a stable, periodic component.  A stellar
flare could be a plausible explanation for such a sudden rise.
However this idea seems incompatible with the sinusoidal emission seen
from $\phi$$\approx$0.5-1.0, and with the fact that flares often
excite iron lines in the blue HIRES orders, which are not observed.
More time-series spectroscopy would be needed to provide a definitive
clarification for this type of non-periodic variability.

Finally, the Li\,\textsc{I} 6708\,\AA\ doublet, which shows no obvious
absorption, as well as the broad K\,\textsc{I} 7699\,\AA\ resonance
line are both shown in Extended Data Figure~\ref{fig:hirescuts}.  In the latter,
narrow telluric absorption features overlap the blue wing of the line.
Neither of these regions shows any notable variability.

@@

Regarding the upper limit, Extended Data Figure~\ref{fig:sed} shows the SED.  While
AllWISE \cite{Cutri2014} yielded a confident W3 detection
(9.8$\sigma$), the W4 extraction yielded only a marginal indication
(1.7$\sigma$) of detectable flux.  Similar to other CPVs
\cite{Stauffer2017,Bouma2024}, the photometric uncertainties from WISE
W1 and W2 allow at most a $\lesssim$2\% excess at 3-5\,$\mu$m relative
to the stellar photosphere, and a $\lesssim$5\% excess at 10\,$\mu$m
(W3).  To estimate the implied mass bound, we assume a dust
temperature $T_{\rm d}$=1500\,K, typical for dust near the star (see
\cite{Zhan2019} for discussion regarding dust sublimation).  We then
treat emission from the dust and star as Planck functions, and require
$L_{\rm d} < f L_\star$, where the factor $f$ is set by the
photometric precision of WISE and $L_{\rm d}$ is the bolometric dust
luminosity.  Given the reported uncertainties, we numerically find
$f<6\cdot$10$^{-3}$.  From the Stefan-Boltzmann law we can then write
$A_{\rm d} < f (T_\star / T_{\rm d})^4 Q_{\rm em}^{-1} (4\pi
R_\star)^2$, for $A_{\rm d}$ the total emitting surface area of the
dust, and $Q_{\rm em}$ an emission efficiency parameter.  Converting
this constraint to a dust mass requires an assumption regarding the
grain properties.  We assume a grain density $\rho_{\rm
d}$=3\,g\,cm$^{-3}$ typical for silicate grains, a fixed grain size
$a$=1\,$\mu$m, and no self-absorption.  This enables the assumption
that $A_{\rm d} = N \pi a^2$, for $N$ the total number of dust grains.
This in turn yields an upper limit on the dust mass of
\begin{equation}
  M_{\rm dust} \lesssim 4 \cdot 10^{17}\, {\rm g}\ 
  \left( \frac{f}{6\cdot10^{-3}} \right)
  \left( \frac{T_\star}{3000\,{\rm K}} \frac{1500\,{\rm K}}{T_{\rm d}} \right)^4
  \left( \frac{Q_{\rm em}}{1} \right)^{-1}
  \left( \frac{R_\star}{0.4\,R_\odot} \right)^2
  \left( \frac{a}{1\,\mathrm{\mu m}} \right)
  \left( \frac{\rho_{\rm d}}{3\,\mathrm{g\,cm^{-3}}} \right).
\end{equation}
